---
output: flexdashboard::flex_dashboard
runtime: shiny
---

```{r setup, include=FALSE}
library(dplyr)
library(flexdashboard)
library(readxl)
library(readr)
library(shiny)
library(rmarkdown)
library(knitr)
library(DT)
library(ggplot2)
library(ggthemes)
library(tidyr)
library(shinythemes)
library(data.table)
library(forecast)
library(rsconnect)
library(plotly)
library(Hmisc)
library(shinythemes)      # Bootswatch color themes for shiny
library(choroplethr)      # Creating Choropleth Maps in R
library(choroplethrMaps)  # Maps used by the choroplethr package

assignInNamespace("cedta.override", c(data.table:::cedta.override,"rmarkdown"), "data.table")

opts_chunk$set(echo = FALSE, comment="", warning = FALSE, message = FALSE, tidy.opts=list(width.cutoff=55), tidy = TRUE)

```

```{r}
visits.national <- fread("https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/flu_view_line(Topright%20%20Chart)_national.csv")

visits.state <- fread("https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/State_Data/ILINet.csv")

visits <- read_csv("https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/flu_view_line(Topright%20%20Chart)_national.csv")

flutype.national <- fread("https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/National_Positive_tests_type.csv")

flutype.state <- fread("https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/State_Data/WHO_NREVSS_Public_Health_Labs.csv",skip=1)

vaccine.national <- fread("https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/usa_vaccination_data.csv")

deaths <- fread("https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/mortality_data.csv")

National.Seasonal.AgebyVirus <- fread(input = "https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/AgeViewBySeason.csv")

National.Weekly.AgebyVirus <- fread(input = "https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/AgeViewByWeek.csv")

dat <- read.csv(url("https://raw.githubusercontent.com/Victoriaxyan/machine-churning/master/Data/State_Data/ILINet.csv"))
```

```{r constants}
region.names <- c("national", unique(visits.state[,REGION]))
age.groups <- names(visits.national)[c(4:6,8:9)]
pred.length <- 4:52 #number of weeks want to plan ahead
datatype <- c("Visit Rates for ILI", 
              names(flutype.national)[c(7,9,10,12,13)],
              age.groups)
Years <- c(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019)
Ages <- c("0-4 yr","5-24 yr","25-64 yr","65+ yr")
Regional.Cut <- c("National", "HHS Regions", "State", "Census")
Seasons <- c("2010-11", "2012-13", "2013-14", "2014-15","2015-16", "2016-17", "2017-18", "2018-19")
select.year<-c("All","2010-2011","2011-2012","2012-2013","2013-2014",
                          "2014-2015","2015-2016","2016-2017","2017-2018",
                          "2018-2019")
Virus <- c("A (Unable to Subtype)","A (H1)","A (H3)","A (H1N1)pdm09","A (Subtyping not Performed)","B (Victoria Lineage)","B (Yamagata Lineage)","B (Lineage Unspecified)","H3N2v")
```

```{r cleaning}
#date format change
visits.national[,MONTH:=lubridate::month(as.Date(paste0(YEAR, "-", 
                                                        WEEK, "-", 10), 
                                                 format = "%Y-%U-%u"))]
visits.national[,date:=as.Date(paste(YEAR,MONTH,1,sep="-"))]
visits.state[,MONTH:=lubridate::month(as.Date(paste0(YEAR, "-",WEEK, "-", 10),
                                              format = "%Y-%U-%u"))]
visits.state[,date:=as.Date(paste(YEAR,MONTH,1,sep="-"))]

visits$month<-lubridate::month(as.Date(paste0(visits$YEAR, "-", visits$WEEK, "-", 10), 
                                       format = "%Y-%U-%u"))
visits$date <- paste(visits$YEAR,visits$month,sep="-")


dat$month<-lubridate::month(as.Date(paste0(dat$YEAR, "-", dat$WEEK, "-", 10), 
                                    format = "%Y-%U-%u"))
dat$date <- paste0(dat$YEAR, "-", dat$month, "-01")		
dat$date <- as.Date(dat$date, "%Y-%m-%d")
dat <- as.data.table(dat)
dat$ILITOTAL <- as.numeric(dat$ILITOTAL)
dat$'TOTAL.PATIENTS' <- as.numeric(dat$'TOTAL.PATIENTS')

data <- dat %>% 
  group_by(REGION, YEAR) %>%                           
  summarise_at(c("ILITOTAL", "TOTAL.PATIENTS"), sum, na.rm = TRUE)

data$ili.percent <- data$ILITOTAL / data$TOTAL.PATIENTS
data$region <- data$REGION
data$region <- sapply(data$region, tolower)
data$REGION <- NULL
data$YEAR <- as.integer(data$YEAR)
data <- as.data.frame(data)

#add the years
visits$duration<-c(rep("2010-2011",52),
                   rep("2011-2012",52),
                   rep("2012-2013",52),
                   rep("2013-2014",52),
                   rep("2014-2015",52),
                   rep("2015-2016",53),
                   rep("2016-2017",52),
                   rep("2017-2018",52),
                   rep("2018-2019",25))

flutype.national[,date:=as.Date(date)]

#add index to vaccine
vaccine.national[,date:=as.Date(paste(year,match(month, month.abb),"1",sep = "-"))]
``` 

```{r functions}

percentage.table <- function(x, digits = 1){
  tab <- table(x)
  percentage.tab <- 100*tab/(sum(tab))
  rounded.tab <- round(x = percentage.tab, digits = digits)
  return(rounded.tab)                           
}


Add.Season.Column <- function(dat){
  i <- 2010
  for(i in Years){
  dat[(Year==i & Week >=40) | (Year==i+1 & Week <=39),season := paste0(i,"-",substring(as.character(i+1),3))]
  i <- i+1
  }
  return(dat)
}

count.table <- function(start,end){
tab <- visits[YEAR>=start & YEAR <= end,.(tot=sum(ILITOTAL)),keyby = YEAR]
return(tab)
}

round.numerics <- function(x, digits){
  if(is.numeric(x)){
    x <- round(x = x, digits = digits)
  }
  return(x)
}
mean.diff <- function(x, y){
  return(mean(x, na.rm=TRUE) - mean(y, na.rm=TRUE))
}
```

```{r AgebyVirus}

National.Seasonal.AgebyVirus <- National.Seasonal.AgebyVirus[Season %in% Seasons,]
National.Weekly.AgebyVirus <- National.Weekly.AgebyVirus[Year %in% Years,]

```


```{r Season}


National.Weekly.AgebyVirus <- Add.Season.Column(National.Weekly.AgebyVirus)

```

```{r Reshape}

National.Weekly.AgebyVirus.long <- melt(National.Weekly.AgebyVirus, 
                                        id.vars = c("Year", "Week", "Age Group","season"),
                                        measure.vars = Virus)

National.Seasonal.AgebyVirus.long <- melt(National.Seasonal.AgebyVirus, 
                                        id.vars = c("Season", "Age Group"),
                                        measure.vars = Virus)

names(National.Weekly.AgebyVirus.long)[names(National.Weekly.AgebyVirus.long) == 'variable'] <- 'Virus'
names(National.Weekly.AgebyVirus.long)[names(National.Weekly.AgebyVirus.long) == 'value'] <- 'Positive Specimens Reported'


names(National.Seasonal.AgebyVirus.long)[names(National.Seasonal.AgebyVirus.long) == 'variable'] <- 'Virus'
names(National.Seasonal.AgebyVirus.long)[names(National.Seasonal.AgebyVirus.long) == 'value'] <- 'Positive Specimens Reported'
```

Introduction
=====================================  

We want to first look at the total flu cases over the past 10 years using a time series plot and identify unusual patterns. For example, the 2018 flu season was the worst since the 2009 swine flu pandemic, and roughly 7.7% of all U.S. citizens seeking medical care experienced flu-like symptoms.

Then the question becomes--is there a way to figure out who was most at risk for contracting a specific type of virus? We plan to break down the outbreak across different states and try to explore the distributions of various flu cases among different age groups. We will try to build interactive plots such as maps, area plots, and bar graphs.

Visit Rates for Influenza-like Illness
=====================================  
Row {data-height=1000}
-------------------------------------
```{r respondents}
#draw the time series plot using visits data
inputPanel(
  selectInput(inputId = "year", label = "Select Year:", 
              choices = select.year, selected = select.year[1]),
  
  checkboxInput(inputId = "age", label = "Breakdown by Age Groups:", value = F)
)

renderPlot({
  
   if(input$year == "All"){
    ili1<-visits%>% 
         mutate(Date = paste(YEAR, month,1,sep = "-"),
                Dates=as.Date(Date,"%Y-%m-%d")) %>%
         group_by(Dates) %>%
         summarise(ilitotal=mean(ILITOTAL,na.rm = T),
                totalp=mean(TOTAL.PATIENTS,na.rm = T))%>%
         mutate(Percentage=round(100*ilitotal/totalp,2))
    
    d=data.frame(date=as.Date('2018-02-1'),event='2018-02-1')

    
    if(input$age==F){

       p1<-ggplot() +
        geom_line(data=ili1, aes(x =Dates, y = Percentage),size=1,col="darkblue")+
        ggtitle("Percentage of Total Reported Influenza Like Illnesses in the US (2010-2019)")+
        labs(caption='Influenza-Like Illness is defined as fever and a cough and/or a sore throat in the absence of a known cause other than flu\nData Source: CDC Weekly US Influenza Surveillance Report')+
        theme_economist_white(base_size = 16)+
        xlab("Year") + ylab("Percentage (%) of ILI patients")+
        theme(plot.title = element_text(hjust=0.5))+
        theme(plot.caption=element_text(hjust=0))+
        scale_x_date(date_breaks = "1 year", date_labels ='%Y')+
        geom_vline(xintercept = as.numeric(as.Date(c('2018-2-1'))),linetype=4,col="blue")+
        geom_text(data=d,mapping=aes(x=date,y=0,label=event),size=5,
            color='blue',angle=90,vjust=1.3,hjust=.1)
    }
    else{
    ili2<-visits%>% 
            mutate(Date = paste(YEAR, month,1,sep = "-"),
            Percentage=ILITOTAL/TOTAL.PATIENTS,
            Dates=as.Date(Date,"%Y-%m-%d")) 
    
    area_dat= ili2 %>%  
          select(Percentage,Dates,YEAR,i=ILITOTAL,young =`AGE.0.4`, 
             young2 = `AGE.5.24`,mid =`AGE.25.49`,old = `AGE.50.64`,old2 = `AGE.65` ) %>% 
          mutate(i=as.numeric(i),young  =as.numeric(young)+as.numeric(young2),
          mid=as.numeric(mid),old =as.numeric(old)+as.numeric(old2) ) %>%
          gather(key=ageGroup,value = ageGroupTotal,c(young,mid,old),factor_key=T)

    area_dat2<-area_dat %>%
         group_by(Dates,ageGroup)%>%
         summarise(ageTotal=sum(ageGroupTotal))

    p1<-ggplot() +
        geom_area(data = area_dat2, aes(x = Dates, y = ageTotal, fill = ageGroup)) +
        scale_fill_brewer(name="Age Groups",labels = c("0-24 years","25-49 years",
                          "50+ years"),palette="YlGn",direction = -1) +
        labs(x="Date",y="Flu Cases by Age Group")+
        theme(plot.caption = element_text(hjust=0))+theme_economist_white(base_size = 16)+
        scale_x_date(date_breaks = "1 year", date_labels ='%Y/%m')
    }
  p1
   }

  else{
    ili1<-visits%>% 
         filter(duration==input$year)%>%
         mutate(Date = paste(YEAR, month,1,sep = "-"),
                date2=as.Date(paste(YEAR,month,WEEK,sep="-"),"%Y-%m-%V"),
         Percentage=round(100*ILITOTAL/TOTAL.PATIENTS,2),
         Dates=as.Date(Date,"%Y-%m-%d"),
         rowN=row_number())
    
  area_dat= ili1 %>%  
    filter(duration==input$year) %>%
    select(Percentage,Dates,WEEK,YEAR,i=ILITOTAL,young =`AGE.0.4`, 
         young2 = `AGE.5.24`,mid =`AGE.25.49`,old = `AGE.50.64`,old2 = `AGE.65` ) %>% 
    mutate(i=as.numeric(i),young  =as.numeric(young)+as.numeric(young2),
         mid=as.numeric(mid),old =as.numeric(old)+as.numeric(old2),
         row=row_number())%>%
  #mutate(Dates = seq.Date(as.Date("2018/01/01"),as.Date("2019/03/01"),by="weeks")) %>%
  gather(key=ageGroup,value = ageGroupTotal,c(young,mid,old),factor_key=T)

  area_dat2<-area_dat %>%
    group_by(ageGroup,Dates)%>%
    summarise(ageTotal=sum(ageGroupTotal))
    
    if(input$age==F){
      p1<-ggplot() +
        geom_line(data=ili1, aes(rowN,  Percentage),size=1,col="darkblue")+
        stat_smooth(color = "#FC4E07", fill = "#FC4E07",method = "loess")+
        ggtitle("Percentage of Weekly Total Reported Influenza Like Illnesses in the US")+
        labs(caption='Influenza-Like Illness is defined as fever and a cough and/or a sore throat in the absence of a known cause other than flu\nData Source: CDC Weekly US Influenza Surveillance Report')+
        theme_economist_white(base_size = 16)+
       ylab("Percentage (%) of ILI patients")+
        theme(plot.title = element_text(hjust=0.5))+
        theme (plot.caption=element_text(hjust=0),
               axis.text.x=element_blank())+
        xlab("Week") 
        #scale_x_continuous(
    #labels = paste("Week",c("40","50","8","18","28","38")," "))
    }
  else{
    p1<-ggplot() +
      geom_area(data = area_dat2, aes(x = Dates, y = ageTotal, fill = ageGroup)) +
      scale_fill_brewer(name="Age Groups",labels = c("0-24 years","25-49 years",
                             "50+ years"),palette="YlGn",direction = -1) +
      labs(x="Date",y="Flu Cases by Age Group")+
      scale_x_date(date_breaks = "1 month")+
      theme(plot.caption = element_text(hjust=0))+
      theme_economist_white(base_size = 14)
  }
p1
  }
   
  
})
```

Virus EDA
===================================

Inputs {.sidebar}
-------------------------------------

```{r AgebyView}
inputPanel(
selectInput(inputId="Seasons", label = "Select Seasons:", choices = Seasons, selected =
Seasons[8]),
checkboxGroupInput(inputId = "Age", label = "Age brackets:", choices=Ages, selected = Ages[1:4]),
checkboxGroupInput(inputId = "Virus", label = "Virus:", choices=Virus, selected = Virus[1:9])
)


```


Column {.tabset}
-------------------------------------
    
### National Virus Distribution - Weekly
    
```{r}

renderPlot({
dat <- National.Weekly.AgebyVirus.long[season %in% input$Seasons & `Age Group` %in% input$Age & Virus %in% input$Virus,]
dat <- dat %>% mutate(row = row_number())
ggplot(dat, aes(reorder(Week,row),`Positive Specimens Reported`, fill = Virus))+geom_bar(stat="identity") + labs(x = "Week") +
theme(axis.text.x = element_text(angle = 90))
})
```


### National Virus Distribution - Seasonal
    
```{r}
renderPlot({
dat <- National.Seasonal.AgebyVirus.long[Season %in% input$Seasons & `Age Group` %in% input$Age & Virus %in% input$Virus,]
dat <- dat[,.(Total=sum(`Positive Specimens Reported`)), by=c("Season","Virus","Age Group")][order(Season)]
ggplot(dat,aes(x="",y=Total,fill=Virus))+geom_bar(stat="identity") + coord_polar("y", start=0)

})
``` 

### Virus Distribution by Age - Weekly
    
```{r}

renderPlot({
dat <- National.Weekly.AgebyVirus.long[season %in% input$Seasons & `Age Group` %in% input$Age & Virus %in% input$Virus,]
dat <- dat %>% mutate(row = row_number())
ggplot(dat, aes(reorder(Week,row),`Positive Specimens Reported`, fill = Virus))+geom_bar(stat="identity") + labs(x = "Week") +
theme(axis.text.x = element_text(angle = 90)) +
facet_grid(~`Age Group`)
})
```


### Virus Distribution by Age - Seasonal
    
```{r}
renderPlot({
dat <- National.Seasonal.AgebyVirus.long[Season %in% input$Seasons & `Age Group` %in% input$Age & Virus %in% input$Virus,]
dat <- dat[,.(Total=sum(`Positive Specimens Reported`)), by=c("Season","Virus","Age Group")][order(Season)]
ggplot(dat,aes(x="",y=Total,fill=Virus))+geom_bar(stat="identity") + coord_polar("y", start=0) +
facet_grid(~`Age Group`)

})
``` 


State Maps for ILI visit rates
=====================================  
Row {data-height=1000}
-------------------------------------

```{r}
ui <- fluidPage(title = 'My First App!',
                theme = shinythemes::shinytheme('flatly'),
                
                sidebarLayout(
                  sidebarPanel(width = 3,
                               sliderInput("years",
                                           "Choose a year:",
                                           min = 2010,
                                           max = 2019,
                                           value = 1,
                                           sep = ""),
                               selectInput("select", 
                                           label = "Select Variable:", 
                                           choices = colnames(data)[2:4], 
                                           selected = 1)),
                  
                  mainPanel(width = 9, 
                            tabsetPanel( 
                              tabPanel(title = 'Output Map', 
                                       plotOutput(outputId = "map")),
                              tabPanel(title = 'Data Table', 
                                       dataTableOutput(outputId = 'table'))))))

server <- function(input, output) {
  
  
  
  output$map <- renderPlot({
    data_20 <- data[data$YEAR == input$years,]
    data_20$value <- data_20[, input$select]
    state_choropleth(data_20,
                     title = input$select,
                     num_colors    = 7)
  })
  
  output$table <- renderDataTable({
    data_20 <- data[data$YEAR == input$years,]
    data_20$value <- data_20[, input$select]
    data_20[order(data_20[input$select]), ]
  })
}

shinyApp(ui = ui, server = server)
```


Models and Forecasts
=====================================  
Row {data-height=500}
-------------------------------------
```{r modeling}

inputPanel(
  #select region
  selectInput(inputId= "Region", label = "Select Region Level:",
              choices = c('',region.names)), #selected = region.names[1]),
  #select data type to predict: visits or any flutype
  selectInput(inputId = "data.type", label = "Select Data",
              choices = datatype), #selected = datatype[1]),

  #select weeks to predict ahead
  numericInput(inputId = "Pred.Length", 
               label = "Number of weeks to predict:", min = 4, max = 104, 
               value = 52, step = 13),
  #select length of original data
  numericInput(inputId = "Orig.Length",
               label = "Number of previous weeks`:", min = 1, max = 445,
               value = 100, step = 5),
  
  helpText("Loading model will take more than 10 seconds, sorry for waiting...")
)

renderPrint({
  if(input$Region==""){print("Please select Region.")}
  else{
  #national level
  if(input$Region=="national"){
    if(input$data.type %in% c(datatype[1], age.groups)){
      if(input$data.type == datatype[1]){
        series <- visits.national[,X..WEIGHTED.ILI] #ILI total rates
      }
      else{
        series <- visits.national[,get(input$data.type)] #ILI by age group
      }
    }
    else{
        series <- flutype.national[,get(input$data.type)] #flu type
    }
  }
  else{
    #state level
    if(input$data.type == datatype[1]){
      series <- visits.state[REGION==input$Region, 
                             na.omit(as.numeric(ILITOTAL)/as.numeric(`TOTAL PATIENTS`))]
      #visits
    }
    else{
      if(input$data.type %in% age.groups){
        series = NULL #no data for state level ILI visits by age groups
      }
      else{
        series = flutype.state[REGION==input$Region,get(input$data.type)] #flu type
      }
    }
  }
  
  if(length(series)<10){
    print("Sorry, the data size is too small to construct models. Please change the Region to National level.")
  }
  else{
    #fitting arima model
    m.auto <- auto.arima(series, max.p = 3) 
    arma <- m.auto$arma #subtracts the orders of SARIMA model
    
    result <- list(`model order` = c(p=arma[1], d=arma[6], q=arma[2],
                                     P=arma[3], D=arma[7], Q=arma[4]),
                   coefficients = data.table(term = names(m.auto$coef),
                                             coefficients = m.auto$coef,
                                             variance = diag(m.auto$var.coef)),
                   criteria = c(AICC = m.auto$aicc, AIC = m.auto$aic,
                                BIC = m.auto$bic)
                   )
    result
  }
}})

renderPlot({
  #national level
  if(input$Region=="") {ggplot()}
  else{
  {if(input$Region=="national"){
    if(input$data.type %in% c(datatype[1], age.groups)){
      if(input$data.type == datatype[1]){
        series <- visits.national[,X..WEIGHTED.ILI] #ILI total rates
      }
      else{
        series <- visits.national[,get(input$data.type)] #ILI by age group
      }
    }
    else{
      series <- flutype.national[,get(input$data.type)] #flu type
    }
  }
  else{
    #state level
    if(input$data.type == datatype[1]){
      series <- visits.state[REGION==input$Region, 
                             na.omit(as.numeric(ILITOTAL)/as.numeric(`TOTAL PATIENTS`))] 
      #visits
    }
    else{
      if(input$data.type %in% age.groups){
        series = rep(0,2) #no data for state level ILI visits by age groups
      }
      else{
        series = flutype.state[REGION==input$Region,get(input$data.type)] #flu type
      }
    }
  }}
  {  
  if(length(series)<10){
    ggplot(data = data.table(week=1:length(series), number=as.numeric(series)),
           aes(x = week, y = number))+
      geom_line(size=2)+
      labs(title = "Data size too small to contruct a model. Please change the region to nation level, or change the data type to 'Visit Rates for ILI'")+
      theme(plot.title = element_text(hjust = 0.5))
  }
  else{
    series <- ts(series, frequency = 52)
    
    #fitting arima model
    m.auto <- auto.arima(series, max.p=3, trace=T) 
    
    #forecast
    pred.list <- forecast(m.auto, h = input$Pred.Length) #point forecasts and CIs
    predictions <- pred.list$mean

    #plot forecast
    orig <- data.table(week = 1:input$Orig.Length,
                       data = series[-1:(-(length(series)-input$Orig.Length))])
    pred <- data.table(week = (input$Orig.Length+1):(input$Orig.Length+
                                                       length(predictions)),
                       data = c(predictions))
    
    ggplot(data = orig, aes(x=week, y=data))+
      geom_line(size = 2, col = "dodgerblue")+
      geom_smooth(data = pred, aes(x=week, y=data, ymax=pred.list$upper[,2],
                                   ymin=pred.list$lower[,2]),
                  col = "LightCoral", size = 2, stat="identity")+
      geom_vline(xintercept = input$Orig.Length,linetype=4,col="brown")+
      geom_text(data = data.frame(week = input$Orig.Length, 
                                  event = "March 2019(12th week)"),
                aes(x = week, y=0, label = event),
                size = 5, col = "Black", angle = 0, vjust=1.3, hjust=.5)+
      labs(title = "Original data with Predictions")+
      theme(plot.title = element_text(hjust = 0.5))
  }}
}})
```

