---
title: "Study on Flu Seasons"
subtitle: "Final Project of GR5243 - Group E"
author: 'Xueyan Zou, Namson Ngo-Le, Rachel Wu, Saloni Mohan'
date: "4/25/2019"
output: ioslides_presentation
---

```{r library, warning=FALSE, echo=FALSE, message=FALSE}
#require(devtools)
#devtools::install_github("cosname/recharts")
library(forecast)
library(data.table)
library(DT)
#library(recharts)
library(ggplot2)
library(dplyr)
library(ggthemes)
library(tidyr)
```

## Introduction to the Problem: Influenza Prediction

**How to predict the intensity of the 2018-2019 Influenza season?  **

Why we are interested:

- In recent years, flu-related deaths have ranged from about 12,000 to — in the worst year — 56,000 (CDC).

- Knowing how to predict -> effective and timely precautionary measures.

## Introduction to the Problem: Influenza Prediction

**Intensity** is decided by the following criteria:

- Number of visits to health care providers for flu

- Number of positive tests on major flu types (H1, H3N2)

Goal: to make predictions on the above criteria for the following groups:

1. Region (National level, State level)

2. Age Groups (Infants, Juniors, Seniors)


## Data Overview: CDC Influenza Data

CDC influenza data in the United States

- National v.s. State

- Patient **visit rates** to health care providers for influenza-like illness (ILI)

- Specimens tested and **the number of positive tests for influenza**

- Vaccination Data

## Data demo: ILI visits, National

```{r echo=FALSE}
visits <- fread("../Data/flu_view_line(Topright  Chart)_national.csv")
visits.cleaned <- visits[,c(2,3,4,5,6,8,9,10)]
visits.cleaned
```

## Data demo: Positive Tests, National

```{r echo=FALSE}
flutype <- fread("../Data/National_Positive_tests_type.csv")
flutype[,c(3,4,5,6,7,9,12)]
```

## Data demo: Vaccination, National

```{r echo=FALSE}
vaccine <- fread("../Data/usa_vaccination_data.csv")
vaccine
```

## Analysis: Statistical Methods

- Model: time series model *SARIMA*

- Reason: The datasets we have are basically time series data. The most widely used model in time series predicting is *SARIMA* models.

- Task: Selecting seasonal, differencing, autoregressive, and moving average parameters.  

- Methodology: Looking at ACF and PACF lags, conduct cross validation and AICC criteria.

## Implementation

Example: visit rates for ILI.  

```{r datacleaning_visits,echo=FALSE}
ili.percent <- visits[YEAR!=2019,.(percent = ILITOTAL/TOTAL.PATIENTS, week = V1)] 
ts.ili.percent <- ts(ili.percent[,percent], frequency = 52)
visits.Apr2019 <- fread("../Data/FluView_LineChart_2019.csv", skip = 1)
ili.2019 <- visits.Apr2019[YEAR==2019, .(percent = ILITOTAL/`TOTAL PATIENTS`)]
```

Choosing paramenters


##Implementation: Fitting model

We have chosen $(2,0,1)\times(0,1,1)_{52}$ as the best model for this data

```{r}
m <- arima(ts.ili.percent, order = c(2,0,1), 
            seasonal = list(order = c(0,1,1), period = 52)) 
m
```

## Implementation: Predicting

```{r}
predictions <- predict(m, n.ahead = 15)$pred
```

```{r echo=FALSE}
par(mfrow=c(2,1))

plot(1:(length(ts.ili.percent) + length(predictions)), 
     c(ts.ili.percent, predictions), type = 'l', col = 1, 
     xlab = "week number starting from 40th week, 2010", 
     ylab = "ILI visits percentage",
     main = "ILI visits percentage real data and predictions")
points((length(ts.ili.percent) + 1) : (length(ts.ili.percent) + length(predictions)),
       ili.2019[,percent], type = 'l', 
       col = "dodgerblue") # real data for 2019 first 15 weeks
points((length(ts.ili.percent) + 1) : (length(ts.ili.percent) + length(predictions)), 
       predictions, type = 'l', col = "lightcoral") #predicted data for 2019 first 15 weeks
legend("topleft", legend = c("real flu visits percent", "predicted flu visits percent"),
       col = c("dodgerblue", "lightcoral"), pch = c("l","l"))

predictions2019 <- predict(m, n.ahead = 52)$pred
plot(1:length(predictions2019), 
     predictions2019, type = 'l', col = "lightcoral", 
     xlab = "week number in 2019", 
     ylab = "ILI visits percentage",
     main = "Predictions of ILI visits percentage in whole 2019")
```

## ILI Visits Overall Data

```{r echo=FALSE, warning=FALSE}
visits[,month:=lubridate::month(as.Date(paste0(visits$YEAR, "-", visits$WEEK, "-", 10), 
                                       format = "%Y-%U-%u"))]
visits[,date:=paste(visits$YEAR,visits$month,sep="-")]

ili1<-visits%>% 
  mutate(Date = paste(YEAR, month,1,sep = "-"),
         Percentage=ILITOTAL/TOTAL.PATIENTS,
          Dates=as.Date(Date,"%Y-%m-%d")) 

d=data.frame(date=as.Date('2018-02-1'),event='2018-02-1')

#PLOT 1: this plot is for the percentage
ggplot() +
  geom_line(data=ili1, aes(x =Dates, y = Percentage))+
  geom_smooth(method='loess',se=F,span=0.5) +
  #ggtitle("Time Plot of the Weekly Total Reported Influenza Like Illnesses in the US (2010-2019)")+
  labs(caption='Figure 1. Influenza-Like Illness is defined as fever and a cough and/or a sore throat in the absence of a known cause other than flu\nData Source: CDC Weekly US Influenza Surveillance Report')+
  theme_economist_white()+
  xlab("Year") + ylab("Percentage of ILI patients")+
  theme(plot.title = element_text(hjust=0.5))+
  theme (plot.caption=element_text(hjust=0))+
  scale_x_date(date_breaks = "1 year", date_labels ='%Y')+
  geom_vline(xintercept = as.numeric(as.Date(c('2018-2-1'))),linetype=4,col="brown")+
  geom_text(data=d,mapping=aes(x=date,y=0,label=event),size=3.5,color='brown',angle=90,vjust=1.3,hjust=.1)
```

## Analysis: ILI visits by Age group

```{r echo=FALSE,warning=FALSE}
ili1$AGE.25.64<-0

area_dat= ili1 %>%  
  filter(YEAR==2018,month>=1) %>%
  select(Percentage,Dates,WEEK,YEAR,i=ILITOTAL,young =`AGE.0.4`, 
         young2 = `AGE.5.24`,mid =`AGE.25.49`,old = `AGE.50.64`,old2 = `AGE.65` ) %>% 
  mutate(i=as.numeric(i),young  =as.numeric(young)+as.numeric(young2),
         mid=as.numeric(mid),old =as.numeric(old)+as.numeric(old2) )%>%
  #mutate(Dates = seq.Date(as.Date("2018/01/01"),as.Date("2019/03/01"),by="weeks")) %>%
  gather(key=ageGroup,value = ageGroupTotal,c(young,mid,old),factor_key=T)

area_dat2<-area_dat %>%
  group_by(Dates,ageGroup)%>%
  summarise(ageTotal=sum(ageGroupTotal))

#area plot by age group
ggplot() +
  geom_area(data = area_dat2, aes(x = Dates, y = ageTotal, fill = ageGroup)) +
  scale_fill_brewer(name="Age Groups",labels = c("0-24 years","25-49 years","50+ years"),palette="YlGn",direction = -1) +
  scale_x_date(date_breaks = "2 weeks")+
  labs(x="Date",y="Flu Cases by Age Group",
       caption = "Figure 3: The total number flu cases broken down by cases\nof patients below the age of 24, between the age 25 and 49, and above the age of 50.\nData Source: CDC Flu View")+
  theme(plot.caption = element_text(hjust=0))+theme_classic()+
  scale_x_date(date_breaks = "1 month", date_labels ='%Y/%m')
```
